Effective collector efficiency of sparse and dense arrays of emergent cylindrical collectors in laminar-turbulent transitional flows

### 2020-08-27 code/analysis.Rmd
done with .pdf for meeting

---
title: "Particle Interception as a Function of Reynolds Number and Collector Density"
author: "Jordan Wingenroth"
date: "8/26/2020"
toc: TRUE
output:
  pdf_document: default
  html_document: default
---

@@

knitr::opts_chunk$set(echo = TRUE)

@@

trap %>%
  group_by(date) %>%
  mutate(mean = mean(sed)) %>%
  arrange(mean) %>%
  ggplot(aes(x = factor(date, levels = unique(date)), y = sed)) +
  geom_boxplot() +
  labs(x = "date", y = "average sediment mass (g)") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

@@

left_join(trap, metadata, by = "date") %>%
  filter(date %in% pumpfinal$date) %>%
  group_by(date) %>%
  mutate(mean = mean(sed)) %>%
  arrange(mean) %>%
  ggplot(aes(x = factor(date, levels = unique(date)), y = sed)) +
  geom_boxplot() +
  labs(x = "date", y = "average sediment mass (g)") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

@@

![Flow Velocity Regression](../pics/frequency_velocity_boxplots.jpeg)

@@

>>>>

---
title: "Particle Interception as a Function of Reynolds Number and Collector Density"
author: "Jordan Wingenroth"
date: "8/26/2020"
toc: TRUE
output:
  pdf_document: default
---

@@

knitr::opts_chunk$set(echo = TRUE, fig.pos = "!H", out.extra = "")

@@

trap %>%
  group_by(date) %>%
  mutate(mean = mean(sed)) %>%
  arrange(mean) %>%
  ggplot(aes(x = factor(date, levels = unique(date)), y = sed)) +
  geom_boxplot() +
  labs(x = "date", y = "sediment mass (mg/trap)") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

@@

trapfinal <- left_join(trap, metadata, by = "date") %>%
  filter(date %in% pumpfinal$date)

trapfinal %>%
  group_by(date) %>%
  mutate(mean = mean(sed)) %>%
  arrange(mean) %>%
  ggplot(aes(x = factor(date, levels = unique(date)), y = sed)) +
  geom_boxplot() +
  labs(x = "date", y = "sediment mass (mg/trap)") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

@@

![](../pics/frequency_velocity_boxplots.jpeg)

It looks like there are a lot of outliers, but that's because there are simply a ton of Vectrino points and the distribution clearly has long tails. See how tiny the IQR boxes are, which by definition encompass 50% of points.

It should also be noted that the back-of-envelope estimate of $v = f / 500$ is extremely close, probably by design.

@@

### Flume volume

The test section is modeled as a rectangular prism, making calculation of volume trivial. However, the water in the flume is in an irregular form. In order to correct for the time water spends flowing outside the test section, where collectors are not acting on particles, we must estimate the total volume of water. We did so using a simple integration of the volumetric flow rate at the drain. We ran an experiment previously in 2018 or 2019, but my physics common sense lapsed: We picked the drain hose up off the ground to take measurements, then stuck it back down in the drain for most of the flow time. This led to an underestimate of about 5-10%. I re-did the analysis. 1st through 4th degree polynomial regression all yield the same volume +/- <0.5%:

```{r}
# flows in ml/s: 
flow <- c(131, 132, 132, 131, 134, 129, 130, 129, 130, 129, 130, 127, 127, 127, 126, 
          123, 126, 125, 124, 125, 127, 125, 123, 122, 125, 124, 118, 123, 125, 125, 
          120, 122, 124, 126, 122, 117, 123, 118, 120, 120, 121, 118, 118, 118, 119, 
          118, 120, 117, 117, 114, 115, 115, 116, 112, 112, 110, 111, 108, 105, 105, 
          102, 100, 100, 102)

# times in fractional hrs of day:  
time <- c(12.13333, 12.21667, 12.28333, 12.41667, 12.5, 12.56667, 12.6, 13.11667,
          13.15, 13.18333, 13.58333, 13.61667, 13.65, 14.25, 14.28333, 14.31667, 
          14.35, 14.38333, 14.68333, 14.71667, 14.75, 14.78333, 14.81667, 14.85, 
          14.88333, 14.93333, 15.35, 15.38333, 15.41667, 15.45, 15.48333, 15.76667, 
          15.8, 15.86667, 15.9, 15.93333, 15.96667, 16.18333, 16.23333, 16.28333, 
          16.31667, 16.33333, 16.36667, 16.55, 16.58333, 16.61667, 16.81667, 16.85, 
          16.9, 16.93333, 16.98333, 17.03333, 17.06667, 17.2, 17.23333, 17.35, 
          17.36667, 17.53333, 17.58333, 17.61667, 17.73333, 17.75, 17.78333, 17.8)

#flow start time:
t_0 <- 11.59166667

#time at which flume was drained to top of test-array holding basin
t_b <- 16.96666667

#time at which basin was drained
t_f <- 17.825

#polynomial model fits:
fits <- lapply(1:4, function(x) {lm(flow ~ poly(time, x, raw = TRUE))})

#finite integral calculator
poly_fin_integr <- function(model,lim1,lim2) {
  sum(coefficients(model)/(1:length(coefficients(model)))*
        lim2^(1:length(coefficients(model))))-
  sum(coefficients(model)/(1:length(coefficients(model)))*
        lim1^(1:length(coefficients(model))))
}

estimates <- lapply(fits, function(x) poly_fin_integr(x, lim1 = t_0, lim2 = t_b))

#convert from mL/s*h to cubic m: 1 mL/s*h = 3600 mL = 3.6 L = 0.0036 cubic m

estimates <- unlist(estimates)*.0036

names(estimates) <- c("linear", "quadratic", "cubic", "4th order")

estimates

I rounded off to 2.43 $\text m^3$. The test section has been measured as $1.95 \times .6 \times .4 =$ `r 1.95*.6*.4` $\text m^3$. Hence the corrective factor is `r 2.43/0.468`.

## Modelling for eta (and all the other vars of interest)

So at this point, we really only need `pumpfinal`, `trapfinal`, maybe `metadata` for good measure, the frequency-to-velocity regression coefficients, and individual constants about the physical setup such as dowel diameter, flume volume, starting sediment mass, dynamic viscosity etc.

Let's clean up for the sake of sanity (if you're reading the .pdf, this is unimportant):

rm(fits, pump, pumpdata, tidypump, tidytrap, trap, trapdata, x, estimates, flow, pumpdate, pumpfiles, t_0, t_b, t_f, time, trapdate, trapfiles, poly_fin_integr)

I'll also mention here that our model is based on the general equation: $$k_{tot} = k_s + k_c + k_{bg}$$ where the k's represent total decay rate and portions of it due to settling, collection, and (background) settling in the rest of the flume outside the test section, respectively.

### k_tot (total time-decay rate)

We estimate k_tot from our pump data by linear regression of `log(mvc)`:

```{r, warning=FALSE}

library(lme4)

fits <- lmList(data = pumpfinal, log(mvc)~t | date)

summary(fits)

The `t` coeffient estimates are the `k_tot` values in units $\text s^{-1}$, no transformation needed aside from a sign change ($\log(\bar{\phi})=\log(\phi_0e^{-kt}) = log(\phi_0)-kt$).

All the t values are high, though the coefficient t-values are relative to 0, meaning they're really just expressing that we're very certain we did indeed add sediment (phew!). But let's see what the fits look like with the data:

cbind(pumpfinal, pred = predict(fits)) %>%
  ggplot() +
  geom_point(aes(x = t, y = mvc), size = .75, alpha = .5) +
  geom_line(aes(x = t, y = exp(pred)), color = "red") +
  facet_grid(pump_freq~dowel_density) +
  scale_x_continuous(breaks = c(0,3000,6000), limits = c(0,6000))

As I noticed back in early 2019, these curves seem to hint at a U-shaped pattern in the residuals.

cbind(pumpfinal, resid = residuals(fits)) %>%
  ggplot(aes(x = t, y = resid)) +
  geom_point() +
  geom_smooth()

I think this is simply an artifact of nonlinear processes not accounted for in the exponential model. As an example of what I mean, let's imagine there were a hidden variable called "stickiness" differentiating walnut shell particles from one another. If by way of accellerated settling due to floculation, or accellerated collection, those particles were disproportionately removed from suspension early in the timespan of the experiment, the average physical properties of the suspended sediment would change over its course. It also seems likely to me that as sediment adheres to collectors, their efficiency decreases on account of their having less open surface area.

Whatever the explanation is, we (and other researchers using the exponential model) assume it is uniform across our independent variables. I don't know how safe that assumption is, but that'll have to wait for a future paper. It might be testable from our data but `n = 12` seems a bit of a small sample. Then again, we've probably done more like `n = 100` runs over the entire history of the project.

Aside from that U-curve, the fit seems reasonably good. Though there are high **absolute** measurements at the beginning, on the **log-scale** they are far less influential than the lower values (`mvc < 10`) at the end. Removing the lower outliers towards the end (`t > 3000`) might be worth trying later if our results don't satisfy.

So then, k_s as a function of our independent vars:

fitdata <- coefficients(fits) %>%
cbind(date = as.numeric(row.names(coefficients(fits)))) %>%
  left_join(metadata, by = "date")

fitdata %>%
  ggplot(aes(x = dowel_density, y = pump_freq, fill = -t, label = round(-t,7))) +
    geom_tile() + 
    geom_text() +
  scale_fill_gradient(low = "red", high = "white")

Mixed feelings. On the one hand it's reassuring that aside from the switcharoo at 20 Hz, all runs with dowels had greater k_t than runs without. However, it seems like there's some serious variance. **What confounders could we be missing???**

### THE PROBLEM IN A NUTSHELL
Having a greater decay rate for the control run than the dowel run, by an amount greater than the difference in $k_s$ between the two runs, implies a negative $k_c$. This seems unrealistic, or at least very out of step with our operating theory. For the rest of this analysis, I'll do what I slipped into my past code and then forgot about: leave out that control run and estimate its values as the mean of the other controls (it does lie at the mean of their independent vars. so weighting the average seems unnecessary). Not ideal! Maybe we should give this set of parameters another shot? But I don't feel great about cherrypicking experimental results that way either.

fitdata[fitdata$date==190802,"t"] <- (fitdata[fitdata$date==190729,"t"] + 
                                        fitdata[fitdata$date==190321,"t"])/2

fitdata
```

### k_s (time-decay due to settling)

$$ m_s = \frac{k_s}{k_{tot}}(1-e^{-kT})m_0 $$

Or, expressed for $k_s$, $$k_s = \frac{m_s}{m_0}(1-e^{-kT})k_{tot}$$

where $m_s$ is mass settled in the test section (estimated from sediment traps), $m_0$ is starting sediment mass (200 g for this study), and $T$ is the total exposure time of the sediment traps (100 min, i.e., 6000 s, for this study).

hist(trapfinal$sed, breaks = 10, density = 10)

There aren't really any clear outliers, as we established in the boxplots above. While splines might be our final tool, for now an average will do.

final <- trapfinal %>%
  group_by(date) %>%
  summarise(m_trap = mean(sed)/1000) %>% #average sediment in trap converted from mg to g
  mutate(m_s = m_trap*1.95*.6/(pi*.0127^2)) %>% #times test section area divided by area of 1 trap
  left_join(fitdata, by = "date") %>%
  mutate(k_s = m_s/200*(1-exp(t*6000))*(-t))

final %>%
  ggplot(aes(x = dowel_density, y = pump_freq, fill = k_s, label = round(k_s,7))) +
    geom_tile() + 
    geom_text() +
  scale_fill_gradient(low = "red", high = "white")

If we're not using the $k_{tot}$ from the 20 Hz control, I think we should probably do the same with k_s, not least because it gives us linear $k_{bg}$ values, though this all still feels a little iffy.

final[final$date==190802,"k_s"] <- (final[final$date==190729,"k_s"] + 
                                      final[final$date==190321,"k_s"])/2

Gladly, settling across the dowel treatments does seem to decrease, with a less clear pattern across flow velocity.

### k_bg (time-decay due to settling in the rest of the flume)

In control runs, $k_c = 0$, so we can estimate the decay rate in the rest of the flume as $k_{bg} = k_{tot}-k_s$

bgvals <- final %>%
  filter(dowel_density=="0000") %>%
  transmute(pump_freq, k_bg = -t - k_s) %>%
  arrange(pump_freq)

bgvals

To add it to the table:

final <- left_join(final, bgvals, by = "pump_freq")

### k_c (time-decay due to collection)

So, now that we have our values in the table, k_c is just a subtraction away:

final <- final %>%
  mutate(k_c = -t - k_s - k_bg)

final %>%
  ggplot(aes(x = dowel_density, y = pump_freq, fill = k_c, label = round(k_c,7))) +
    geom_tile() + 
    geom_text() +
  scale_fill_gradient(low = "red", high = "white")

final %>%
  ggplot(aes(x = as.numeric(pump_freq), y = k_c, color = dowel_density)) +
  geom_line()

Unlike the calculation with the measured values at the 20 Hz control, all k_c are in the green, but the pattern is still quite odd.

### eta

#temp in flume measured at 22.2C with a calibrated thermometer
#plugged into https://www.engineeringtoolbox.com/water-dynamic-kinematic-viscosity-d_596.html
visc = 9.509e-7 #kinematic viscosity, m2/s

d = .003175 #dowel diameter

final <- final %>%
  transmute(frontal_area = as.numeric(dowel_density)*.003175, 
            u = as.numeric(pump_freq)/500, #velocity 
            Re = u*.003175/visc, #Reynolds #
            k_t = -t, 
            k_c, k_s, k_bg) %>%
  mutate(eta = k_c/u/frontal_area) %>%
  mutate(eta = eta * 2.43/(1.95*.4*.6)) # don't forget to correct for time out of test section!

final %>%
  filter(frontal_area != 0) %>%
  ggplot(aes(x = Re, y = eta, color = factor(frontal_area))) +
  geom_line() +
  geom_point()

## Possible additions

- Turbulence data
    - The calculations are done 
    - But it might be good to include here for the sake of comprehensiveness

- i did find a way to propogate error through these calculations quite easily
    - I think including error bars on our main graph (eta over Rey) might be a good idea?
    - Is it irresponsible to do so with `n = 1` per treatment?

### 2020-08-27 drafts/outline.tex
Added eta figure to manuscript
done with .pdf for meeting

\begin{figure}[htbp]
\includegraphics[width=15cm]{pics/flume_with_sedtraps.png}
\centering
\caption{Conceptual diagram of the laboratory flume. Labeled parts are: 1) pump, 2) magnetic flowmeter, 3) test section, and 4) honeycomb flow collimator. Arrows indicate direction of flow. A side view of the open-channel part of the system is included. Some dimensions not to scale. Red circles in the top-down view and black rectangles in the side view of the test section show the locations of the sediment traps.}
\label{fig:floorplan}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=6in]{etafig.png}
\centering
\caption{A plot showing our calculated effective collector efficiency ($\eta^\prime$) values for various values of $I_c$ and $Rey_c$}
\label{fig:vectrino}
\end{figure}

>>>>

\begin{figure}[htbp]
\includegraphics[width=15cm]{../pics/flume_with_sedtraps.png}
\centering
\caption{Conceptual diagram of the laboratory flume. Labeled parts are: 1) pump, 2) magnetic flowmeter, 3) test section, and 4) honeycomb flow collimator. Arrows indicate direction of flow. A side view of the open-channel part of the system is included. Some dimensions not to scale. Red circles in the top-down view and black rectangles in the side view of the test section show the locations of the sediment traps.}
\label{fig:floorplan}
\end{figure}

@@

\begin{figure}[htbp]
\includegraphics[width=6in]{etafig.png}
\centering
\caption{A plot showing our calculated effective collector efficiency ($\eta^\prime$) values for various values of frontal area ($I_cd_c$) and $\Rey_c$}
\label{fig:eta}
\end{figure}
